Modelling at-bats in baseball using the generalised Pareto distribution

### 2023-04-27 bibat/examples/baseball/docs/report.qmd
Update vignette

---
title: "Modelling at-bats in baseball using the generalised Pareto distribution"
subtitle: "A report created using bibat version 0.0.12"
author: "Teddy Groves"
bibliography: bibliography.bib
reference-location: margin
toc: true
format:
  html:
    embed-resources: true
filters:
  - include-code-files
---

@@

```zsh
$ pip install bibat
```

Next I started bibat's wizard like this:

```zsh
bibat
```

@@

```zsh
baseball
├── CODE_OF_CONDUCT.md
├── LICENSE
├── Makefile
├── README.md
├── bibat_version.txt
├── data
│   └── raw
│       ├── raw_measurements.csv
│       └── readme.md
├── docs
│   ├── bibliography.bib
│   ├── img
│   │   ├── example.png
│   │   └── readme.md
│   └── report.qmd
├── inferences
│   ├── fake_interaction
│   │   └── config.toml
│   ├── interaction
│   │   └── config.toml
│   └── no_interaction
│       └── config.toml
├── investigate.ipynb
├── plots
│   ├── posterior_ll_comparison.png
│   └── posterior_predictive_comparison.png
├── prepare_data.py
├── pyproject.toml
├── requirements-tooling.txt
├── requirements.txt
├── sample.py
├── src
│   ├── __init__.py
│   ├── data_preparation_functions.py
│   ├── inference_configuration.py
│   ├── prepared_data.py
│   ├── readme.md
│   ├── stan
│   │   ├── custom_functions.stan
│   │   ├── multilevel-linear-regression.stan
│   │   └── readme.md
│   ├── stan_input_functions.py
│   └── util.py
└── tests
    ├── test_integration
    │   └── test_data_preparation.py
    └── test_unit
        ├── test_inference_configuration.py
        └── test_util.py
```

@@

Some cogs turned, some new lines appeared in my terminal and some new files were
created - nice, the setup worked!

@@

To fetch raw data from the internet, I wrote a new script called `fetch_data.py`:

```{.python include="../fetch_data.py"}
```

@@

```zsh
$ source .venv/bin/activate
$ python fetch_data.py
```

Since this worked, I added a new makefile target for the raw data files:

```makefile
RAW_DATA = data/raw/2006.csv \
	data/raw/bdb-main.csv \
	data/raw/bdb-post.csv \
	data/raw/bdb-apps.csv

...
$(RAW_DATA):
	. $(ACTIVATE_VENV) && python fetch_data.py
```

@@

```zsh
$ rm data/raw/raw_measurements.csv
```

# Preparing the data
The first step in preparing data is to decide what prepared data looks like for
the purposes of our analysis. Bibat provides dataclasses called `PreparedData`
and `MeasurementsDF ` in the file `src/prepared_data.py` which can help get us
started with this.

As it happens, prepared data looks very similar in our analysis and the example.
All we need to do is change the `MeasurementsDF` definition a little^[note that
this class uses [pandera](https://pandera.readthedocs.io), a handy library for
defining what a pandas dataframe should look like]:

```python
from typing import Optional
import pandera as pa

# ...

class MeasurementsDF(pa.SchemaModel):
    """A PreparedData should have a measurements dataframe like this.

@@

    player_season: pa.typing.Series[str]
    season: pa.typing.Series[str]
    n_attempt: pa.typing.Series[int] = pa.Field(ge = 1)
    n_success: pa.typing.Series[int] = pa.Field(ge = 0)
```

Next we can write some functions that create `PreparedData` objects. These
should live in the file `data_preparation_functions.py`. In this case we write a
couple of data preparation functions: `prepare_data_2006` and
`prepare_data_bdb`:

```{.python include="../src/data_preparation_functions.py"}
```

Finally we need to update `prepare_data.py`, the script that runs the data
preparation functions. Again there isn't much to change from the example
analysis.

```{.python include="../prepare_data.py"}
```

To check that all this works, we can run the script `prepare_data.py` manually
or using `make analysis`. Now if we look in the file
`data/prepared/bdb/measurements.csv` we should see some lines that look like
this:

@@

```{stan}
real gpareto_lpdf(vector y, real ymin, real k, real sigma) {
  // generalised Pareto log pdf 
  int N = rows(y);
  real inv_k = inv(k);
  if (k<0 && max(y-ymin)/sigma > -inv_k)
    reject("k<0 and max(y-ymin)/sigma > -1/k; found k, sigma =", k, ", ", sigma);
  if (sigma<=0)
    reject("sigma<=0; found sigma =", sigma);
  if (fabs(k) > 1e-15)
    return -(1+inv_k)*sum(log1p((y-ymin) * (k/sigma))) -N*log(sigma);
  else
    return -sum(y-ymin)/sigma -N*log(sigma); // limit k->0
}
```

@@

```{.stan include="../src/stan/gpareto.stan"}
```

@@

```{.stan include="../src/stan/normal.stan"}
```

# Generating Stan inputs
Next we need to tell our analysis how to turn some prepared data into a
dictionary that can be used as input for Stan. Bibat assumes that this task is
handled by functions that live in the file `src/stan_input_functions.py`, each
of which takes in a `PreparedData` and returns a Python dictionary. You can
write as many Stan input functions as you like and choose which one to run for
any given inference.

We can start by defining some Stan input functions that pass arbitary prepared data
on to each of the models:

```{.python include="../src/stan_input_functions.py" start-line=12 end-line=34}
```

But why stop there? It can also be useful to generate Stan inputs consistently
with a model, based on some hardcoded hyperparameter values. Here are some
functions that do this for both of our models:

```{.python include="../src/stan_input_functions.py" start-line=37}
```

# Specifying inferences
Now all the building blocks for making statistical inferences - raw data, data
preparation rules, statistical models and recipes for turning prepared data into
model inputs - are in place. The last step before actually running Stan is to
write down how put these blocks together. Bibat has another concept for this,
called 'inferences'.

@@

```zsh
.
├── gpareto2006
│   └── config.toml
└── normal2006
    └── config.toml
```

@@

# Investigating the inferences
Now that the inferences are ready it's time to check them out. Bibat provides a
jupyter notebook called `investigate.ipynb` for exactly this purpose.

@@

After running `make analysis` one more time, I went back to the notebook
`investigate.ipynb` and made plots of both models' posterior 1%-99% success rate
intervals for both datasets:

>>>>

---
title: "Modelling at-bats in baseball using the generalised Pareto distribution"
subtitle: "A report created using bibat version 0.1.3"
author: "Teddy Groves"
bibliography: bibliography.bib
reference-location: margin
toc: true
format:
  html:
    embed-resources: true
filters:
  - include-code-files
---

@@

```zsh
> pip install bibat
```

Next I ran bibat's wizard like this:

```{.zsh}
> bibat
Welcome to the Batteries-Included Bayesian Analysis Template!
What is your project called? [Project Name]: Baseball
What should the project repository be called [baseball]: baseball
What is your name? [Author name]: Teddy Groves
What is your email? [author@email.com]:
Who should be the code of conduct contact? [author@email.com]:
Please briefly describe your project [A short description of the project.]: Comparison of distributions for modelling baseball hitting
Choose an open source license from these options: (MIT, BSD-3-Clause, No license file) [MIT]:
How would you like to document your project? (Quarto, Sphinx, No docs) [Quarto]:
Would you like to create a tests directory? [y]: n
Would you like to create a .github directory? [y]: n
> 
```

@@

```{.zsh}
baseball
├── CODE_OF_CONDUCT.md
├── LICENSE
├── Makefile
├── README.md
├── baseball
│   ├── __init__.py
│   ├── data_preparation.py
│   ├── inference_configuration.py
│   ├── investigate.ipynb
│   ├── notebooks
│   ├── prepare_data.py
│   ├── readme.md
│   ├── sample.py
│   ├── stan
│   │   ├── custom_functions.stan
│   │   ├── multilevel-linear-regression.stan
│   │   └── readme.md
│   ├── stan_input_functions.py
│   └── util.py
├── bibat_version.txt
├── data
│   └── raw
│       ├── raw_measurements.csv
│       └── readme.md
├── docs
│   ├── bibliography.bib
│   ├── img
│   │   ├── example.png
│   │   └── readme.md
│   └── report.qmd
├── inferences
│   ├── fake_interaction
│   │   └── config.toml
│   ├── interaction
│   │   └── config.toml
│   └── no_interaction
│       └── config.toml
├── plots
│   ├── posterior_ll_comparison.png
│   └── posterior_predictive_comparison.png
├── pyproject.toml
└── requirements.txt
```

@@

This ran without errors, running some data preparation functions and then analysing the data with Stan in a few configurations.

@@

To fetch raw data from the internet, I wrote a new script in the file
`baseball/fetch_data.py`:

```{.python include="../baseball/fetch_data.py"}
```

@@

```zsh
> source .venv/bin/activate
(baseball) > python baseball/fetch_data.py
Fetching 2006 data from https://raw.githubusercontent.com/stan-dev/example-models/master/knitr/pool-binary-trials/baseball-hits-2006.csv
Writing 2006 data to data/raw/2006.csv
Fetching bdb-main data from https://raw.githubusercontent.com/chadwickbureau/baseballdatabank/master/core/Batting.csv
Writing bdb-main data to data/raw/bdb-main.csv
Fetching bdb-post data from https://raw.githubusercontent.com/chadwickbureau/baseballdatabank/master/core/BattingPost.csv
Writing bdb-post data to data/raw/bdb-post.csv
Fetching bdb-apps data from https://raw.githubusercontent.com/chadwickbureau/baseballdatabank/master/core/Appearances.csv
Writing bdb-apps data to data/raw/bdb-apps.csv
```

@@

```zsh
(baseball) > rm data/raw/raw_measurements.csv
```

# Preparing the data

The first step in preparing data is to decide what prepared data looks like for
the purposes of our analysis. Bibat provides dataclasses called `PreparedData`
and `MeasurementsDF` to help get started with this, which I found in the file
`baseball/prepared_data.py`.

As it happens, prepared data looks very similar in this analysis and the
example. All I had to do was change the `MeasurementsDF` definition a
little^[note that this class uses [pandera](https://pandera.readthedocs.io), a
handy library for defining what a pandas dataframe should look like]:

```python
class MeasurementsDF(pa.SchemaModel):
    """A PreparedData should have a measurements dataframe like this.

@@

    player_season: Series[str]
    season: Series[str] = pa.Field(coerce=True)
    n_attempt: Series[int] = pa.Field(ge = 1)
    n_success: Series[int] = pa.Field(ge = 0)
```

The next step is to write functions that return `PreparedData` objects. In this
case I wrote a couple of data preparation functions: `prepare_data_2006` and
`prepare_data_bdb`:

```{.python include="../baseball/data_preparation.py" start-line=125}
```

To check that all this worked, I ran the data preparation script manually:^[I
could also have just run `make analysis` again]

```{.zsh}
> source .venv/bin/activate
(baseball) > python baseball/prepare_data.py
```

Now the folder `data/prepared/bdb` contained a file
`data/prepared/bdb/measurements.csv` that looked like this:

@@

```{.stan}
real gpareto_lpdf(vector y, real ymin, real k, real sigma) {
  // generalised Pareto log pdf 
  int N = rows(y);
  real inv_k = inv(k);
  if (k<0 && max(y-ymin)/sigma > -inv_k)
    reject("k<0 and max(y-ymin)/sigma > -1/k; found k, sigma =", k, ", ", sigma);
  if (sigma<=0)
    reject("sigma<=0; found sigma =", sigma);
  if (fabs(k) > 1e-15)
    return -(1+inv_k)*sum(log1p((y-ymin) * (k/sigma))) -N*log(sigma);
  else
    return -sum(y-ymin)/sigma -N*log(sigma); // limit k->0
}
```

@@

```{.stan include="../baseball/stan/gpareto.stan"}
```

@@

```{.stan include="../baseball/stan/normal.stan"}
```

# Generating Stan inputs

Next I needed to tell the analysis how to turn some prepared data into a
dictionary that can be used as input for Stan. Bibat assumes that this task is
handled by functions that live in the file `baseball/stan_input_functions.py`,
each of which takes in a `PreparedData` and returns a Python dictionary. You
can write as many Stan input functions as you like and choose which one to run
for any given inference.

I started by defining some Stan input functions that pass arbitary prepared
data on to each of the models:

```{.python include="../baseball/stan_input_functions.py" start-line=11 end-line=34}
```

But why stop there? It can also be useful to generate Stan input data from a
model with a model, by running it in simulation model with hardcoded parameter
values. Here are some functions that do this for both of our models:

```{.python include="../baseball/stan_input_functions.py" start-line=36}
```

# Specifying inferences

Now all the building blocks for making statistical inferences - raw data, data
preparation rules, statistical models and recipes for turning prepared data
into model inputs - were in place. The last step before actually running Stan
was to write down how put the blocks together. Bibat has another concept for
this, called 'inferences'.

@@

```zsh
inferences
├── gpareto2006
│   └── config.toml
└── normal2006
    └── config.toml
```

@@

# Investigating the inferences

Now that the inferences are ready it's time to check them out. Bibat provides a
jupyter notebook at `baseball/investigate.ipynb` for exactly this purpose.

@@

After running `make analysis` one more time, I went back to the notebook
`baseball/investigate.ipynb` and made plots of both models' posterior 1%-99%
success rate intervals for both datasets:
